Note on Linear Biases in MORUS

### 2018-04-08 morus.tex
more writing

\usepackage{amsmath,amssymb}
\usepackage{ifthen}
\usepackage{subcaption}
\usepackage{booktabs}
\usepackage{siunitx}
\usepackage[section]{placeins}
\usepackage[hidelinks]{hyperref}

@@

We will also use the following notation, where an \emph{encryption step} refers to one call to the \StateUpdate{} function:\\
\centerline{
\def\arraystretch{1.25}
\begin{tabular}{lcl}
$C^t$ &:& the ciphertext block output during the $t$-th encryption step.\\
$C^t_j$ &:& the $j$-th bit of $C^t$, with $C^t_0$ being the rightmost bit.\\
$S^t_i$ &:& the $i$-th register at the beginning of $t$-th encryption step.\\
$S^t_{i,j}$ &:& the $j$-th bit of $S^t_i$, with $S^t_{i,0}$ being the rightmost bit.
\end{tabular}
}
In the remainder, the $0$-th encryption step will denote the encryption step where our linear trail starts. Any encryption step could be chosen for that purpose, as long as at least four more encryption steps follow. In particular the $0$-th encryption step from the perspective of the trail does not have to be the first encryption step after initialization.

@@

\section{A Linear Bias in \cipher{MORUS}}

To simplify the description of the attack, we assume all plaintext blocks are zero. However the same results would hold as long as the plaintext is fixed. Let $w$ denote the word size, i.e. 32 for \cipher{MORUS640}, and 64 for \cipher{MORUS1280}. Recall that the inner state of the cipher consists of five $4w$-bit registers $S_0,\dots,S_4$, each containing four $w$-bit words.

@@

We begin with a few observations about the \StateUpdate{} function. Beside XOR and AND operations, the \StateUpdate{} function uses two types of bit rotations:
\begin{enumerate}
\item \emph{word-wise} rotations perform a circular shift on each word within a register;
\item \emph{register-wise} rotations perform a circular shift on a whole register.
\end{enumerate}
The second type of rotation always shifts registers by a multiple of the word size (i.e. 32 for \cipher{MORUS640}). This amounts to a (circular) permutation of the words within the register: for example, if a register contains the words $(A, B, C, D)$, and a register-wise rotation by 32 bits to the left is performed, then the register now contains the words $(B, C, D, A)$.

Since we want to build a linear trail, we are interested in linear combinations of bits within a register.
\begin{definition}
Let $w$ denote the word size in bits, and $r = 4w$ denote the register size. A linear combination of the form:
\[
S^t_{i,j(0)} \oplus S^t_{i,j(1)} \oplus \dots \oplus S^t_{i,j(k)}
\]
is said to be \emph{symmetric} iff the set of bits $S^t_{i,j(0)}, \dots, S^t_{i,j(k)}$ is left invariant by a circular shift by $w$ bits; that is, iff:
\[
\{j(i) : i\leq k\} = \{j(i) + w \text{\rm{} mod } r : i\leq k\}.
\]
\end{definition}
\emph{Example.} The following linear combination is symmetric for \cipher{MORUS640}, i.e. $w = 32$:
\begin{equation}
S^t_{0,0} \oplus S^t_{0,32} \oplus S^t_{0,64} \oplus S^t_{0,96}.
\label{eq:symmetric}
\end{equation}

@@

Because the trail we are going to build is relatively complex, we will first describe it on \cipher{MiniMORUS640}. We will then extend it to the full \cipher{MORUS640} via the previous symmetry. The trail for \cipher{MORUS1280} will be essentially the same.

@@

\section{Linear Trail for \cipher{MiniMORUS640}}

@@

The trail fragments are depicted on \autoref{fig:minimorusfragments}, and may be represented as follows (recall that the $b_i$'s are rotation constants):
\begin{alignat*}{2}
\alpha^t_i:\quad&& C^t_i &\to S^{t+1}_{0,i+b_0} \\
\beta^t_i:\quad&& C^t_i, S^t_{0,i} &\to S^t_{1,i} \\
\gamma^t_i:\quad&& S^t_{1,i}, S^{t+1}_{1,i+b_1} &\to S^t_{4,i} \\
\delta^t_i:\quad&& S^t_{4,i}, S^{t+1}_{4,i+b_4} &\to S^{t+1}_{2,i} \\
\varepsilon^t_i:\quad&& S^t_{2,i}, S^{t+1}_{2,i+b_2} &\to S^{t+1}_{0,i}.
\end{alignat*}

In all cases except $\alpha^t_i$, the trail fragment approximates a single AND gate by zero, which holds with probability 3/4, and hence the trail fragment has weight 1. In the case of $\alpha^t_i$, two AND gates are involved; however the two gates share an entry in common, so their sum can be written as:
\[
x \cdot y \oplus x \cdot z = x \cdot (y \oplus z).
\]
As a result, the trail fragment $\alpha^t_i$ also has a weight of 1. Another way to look at this phenomenon is that the trail holds for two different approximations of the AND gates: the alternative approximation is depicted by a dashed line on \autoref{fig:minimorusfragments}.

@@

\subsection{Trail Equation}

@@

In order to build the equation for the full trail, we start with $E^3_0$:
\[
S^3_{2,0} \oplus S^{4}_{2,b_0} \oplus S^{4}_{0,0} = S^3_{3,0} \cdot S^t_{4,0}.
\]
In order to cancel the $S^{4}_{0,0}$ term on the left-hand side, we add in the equation $A^3_{-b_0}$ (where the sum of two equations of the form $a = b$ and $c = d$ is defined to be $a+c = b+d$). This yields:
\begin{align*}
&S^3_{2,0} \oplus S^{4}_{2,7} \oplus C^{3}_{0,-b_0}\\
=\; &S^3_{3,0} \cdot S^t_{4,0} \oplus S^t_{1,-b_0} \oplus S^t_{3,-b_0} \oplus S^3_{1,-b_0} \cdot S^3_{2,-b_0} \oplus S^3_{2,-b_0} \cdot S^t_{3,-b_0}.
\end{align*}
We then need to cancel two terms of the form $S^t_{2,i}$. To do this, we add in the equations $D^t_i$ for appropriate choices of $t$ and $i$. This replaces the two $S^t_{2,i}$ terms by four $S^t_{4,i}$ terms. By using equation $B^t_i$ four times, we can then replace these four $S^t_{4,i}$ terms by eight $S^t_{1,i}$ terms. By applying equation $B^t_i$ eight times, these eight $S^t_{1,i}$ terms can in turn be replaced by eight $S^t_{0,i}$ terms (and some ciphertext terms). Finally, applying $A^t_i$ eight times allows us to replace these eight $S^t_{0,i}$ terms by only ciphertext bits. We do not write the whole process as it is rather tedious. Ultimately, for \cipher{MORUS640}, this yields the equation:

\allowdisplaybreaks
\begin{flalign*}
&&&C^0_{27} \oplus C^1_{0} \oplus C^1_{2} \oplus C^1_{8} \oplus C^1_{26} \oplus C^2_{1} \oplus C^2_{13} \oplus C^2_{15}&&\\
&&\oplus\; &C^2_{27} \oplus C^2_{31} \oplus C^3_{6} \oplus C^3_{12} \oplus C^3_{14} \oplus C^3_{20} \oplus C^4_{19}&&\\
&&=\; &S^0_{1,27} \oplus S^0_{3,27} \oplus S^0_{1,27} \cdot S^0_{2,27} \oplus S^0_{2,27} \cdot S^0_{3,27}&&\text{weight 1}\\
&&\oplus\; &S^1_{1,26} \oplus S^1_{3,26} \oplus S^1_{1,26} \cdot S^1_{2,26} \oplus S^1_{2,26} \cdot S^1_{3,26}&&\text{weight 1}\\
&&\oplus\; &S^1_{1,8} \oplus S^1_{3,8} \oplus S^1_{1,8} \cdot S^1_{2,8} \oplus S^1_{2,8} \cdot S^1_{3,8}&&\text{weight 1}\\
&&\oplus\; &S^2_{3,1} \oplus S^2_{1,1} \oplus S^2_{1,1} \cdot S^2_{2,1} \oplus S^2_{2,1} \cdot S^2_{3,1}&&\text{weight 1}\\
&&\oplus\; &S^2_{1,7} \oplus S^2_{3,7} \oplus S^2_{1,7} \cdot S^2_{2,7} \oplus S^2_{2,7} \cdot S^2_{3,7}&&\text{weight 1}\\
&&\oplus\; &S^2_{1,15} \oplus S^2_{3,15} \oplus S^2_{1,15} \cdot S^2_{2,15} \oplus S^2_{2,15} \cdot S^2_{3,15}&&\text{weight 1}\\
&&\oplus\; &S^2_{1,27} \oplus S^2_{3,27} \oplus S^2_{1,27} \cdot S^2_{2,27} \oplus S^2_{2,27} \cdot S^2_{3,27}&&\text{weight 1}\\
&&\oplus\; &S^1_{1,2} \oplus S^1_{3,2} \oplus S^1_{1,2} \cdot S^1_{2,2} \oplus S^1_{2,2} \cdot S^1_{3,2}&&\text{weight 1}\\
&&\oplus\; &S^3_{3,14} \oplus S^3_{1,14} \oplus S^3_{1,14} \cdot S^3_{2,14} \oplus S^3_{2,14} \cdot S^3_{3,14}&&\text{weight 1}\\
&&\oplus\; &S^2_{0,0} \cdot S^2_{1,0}&&\text{weight 1}\\
&&\oplus\; &S^2_{2,31} \cdot S^2_{3,31}&&\text{weight 1}\\
&&\oplus\; &S^2_{3,0} \cdot S^2_{4,0}&&\text{weight 1}\\
&&\oplus\; &S^3_{0,7} \cdot S^3_{1,7}&&\text{weight 1}\\
&&\oplus\; &S^3_{2,6} \cdot S^3_{3,6}&&\text{weight 1}\\
&&\oplus\; &S^3_{2,12} \cdot S^3_{3,12}&&\text{weight 1}\\
&&\oplus\; &S^4_{2,19} \cdot S^4_{3,19}.&&\text{weight 1}
\end{flalign*}

\subsection{Bias of the Trail}

In the equation from the previous section, each line on the right-hand side of the equality involves distinct terms, and each line has a weight of 1. By the piling-up lemma, it follows that if we assume distinct $S^t_{i,j}$ terms to be uniform and independent, then the expression on the right-hand side has a weight of 16. Hence the linear combination of ciphertext bits on the left-hand side has a bias of $2^{-16}$.

@@

To confirm that our analysis is correct, we ran experiments on an implementation of \cipher{MiniMORUS640} and \cipher{MiniMORUS1280}. In both cases, we measured the bias of two halves $\chi_1$ and $\chi_2$ of the full trail, as well as the bias of the full trail itself (denoted by $\chi$). Results are displayed on \autoref{tab:miniapproximations}. 
While our analysis from the previous section predicts a bias of $2^{-16}$, experiments confirm a slightly higher bias of $2^{-15.5}$ for \cipher{MORUS640}. The discrepancy of $2^{-0.5}$ probably arises from the fact that register bits across different rounds are not completely independent.

@@

\begin{figure}
  \substatesfalse
  % \substatesfalse to label state words and/or masks
  \centering
  \begin{subfigure}{.32\textwidth}
  \centering
  \begin{tikzpicture}[xscale=0.75,yscale=1.5]%{{{
    \printstate
    \draw[trail, beta]
      (C) -- node[right] {$i$} (lll-1)
      (lll-1) -- (tlll-1) (lll-1) -- (xor-1) (xor-1) -- (xnd-1)
      (tlll-1) -- (W-20) node[above] {$i$}
      (xor-1) -- (txor-1) %(txor-1) -- (W-21) node[above] {$i$}
      %(xnd-1) -- (and-1)
      (W40) node[below] {\phantom{$i$}}
      ;
    \draw[trail, gamma]
      %(W-21) node[above] {$i$}
      (txor-1) -- (xnd1)
      (xnd1) -- (xor1) (xor1) -- (lll1)
      %(xnd1) -- (and1)
      (xor1) -- (txor1)
      (txor1) -- (W-24) node[above] {$i$}
      (lll1) -- (W41) node[below] {$i+b_1$}
      ;
    %\draw[trail, beta,  dotted] (and-1) -- (and-1-|and1);
    %\draw[trail, gamma, dotted] (and1)  -- (and-1-|and1) node[above, black] {=};
    \draw[trail, beta]  (xnd-1) -- ++(-.5,0) coordinate (tmp) -- (tmp|-xnd1);
    \draw[trail, gamma] (tmp|-xnd1) node[below, black] {=} -- (xnd1);
  \end{tikzpicture}%}}}
  \caption*{\cipher{MiniMORUS}: weight 0 (not 2)} %$S^{j,1}_i, S^{j,4}_i, S^{j+1,1}_{i-1}$ ($w\!=\!1$)}
  \end{subfigure}
  \qquad
  \begin{subfigure}{.32\textwidth}
  \centering
  \begin{tikzpicture}[xscale=0.75,yscale=1.5]%{{{
    \printstate
    \draw[trail, beta]
      (C) -- node[right] {$i$} (lll-1)
      (lll-1) -- (tlll-1) (lll-1) -- (xor-1) (xor-1) -- (xnd-1)
      (tlll-1) -- (W-20) node[above] {$i$}
      (xor-1) -- (txor-1) %(txor-1) -- (W-21) node[above] {$i$}
      (xnd-1) -- (and-1)
      (W40) node[below] {\phantom{$i$}}
      ;
    \draw[trail, gamma]
      %(W-21) node[above] {$i$}
      (txor-1) -- (xnd1)
      (xnd1) -- (xor1) (xor1) -- (lll1)
      (xnd1) -- (and1)
      (xor1) -- (txor1)
      (txor1) -- (W-24) node[above] {$i$}
      (lll1) -- (W41) node[below] {$i+b_1$}
      ;
    \draw[trail, beta,  dashed] (and-1.east|-tanB-1) -- (tanB-1);
    \draw[trail, gamma, dashed] (tanB-1)  -- (tanB1) -- (tanB1-|and1.east);
  \end{tikzpicture}%}}}
  \caption*{\cipher{MORUS}: weight $4 \times 1$ (not $4 \times 2$)} %$S^{j,1}_i, S^{j,4}_i, S^{j+1,1}_{i-1}$ ($w\!=\!1$)}
  \end{subfigure}
  \caption{Weight of $\beta_i \oplus \gamma_i$ for \cipher{MiniMORUS} and \cipher{MORUS}.}
  \label{fig:minimorusfragments}
\end{figure}

>>>>

\usepackage{amsmath,amssymb}
\usepackage{ifthen}
\usepackage{subcaption}
\usepackage{booktabs}
\usepackage{siunitx}
\usepackage[section]{placeins}
\usepackage[hidelinks]{hyperref}
\usepackage{cleveref}

\allowdisplaybreaks%this is to allow page breaks within multiline equations (to accomodate full trail equations)

@@

We will also use the following notation, where an \emph{encryption step} refers to one call to the \StateUpdate{} function:\\
\centerline{
\def\arraystretch{1.25}
\begin{tabular}{lcl}
$C^t$ &:& the ciphertext block output during the $t$-th encryption step.\\
$C^t_j$ &:& the $j$-th bit of $C^t$, with $C^t_0$ being the rightmost bit.\\
$S^t_i$ &:& the $i$-th register at the beginning of $t$-th encryption step.\\
$S^t_{i,j}$ &:& the $j$-th bit of $S^t_i$, with $S^t_{i,0}$ being the rightmost bit.
\end{tabular}
}
In the above notation, bit positions are always taken modulo the register size, i.e. 128 for \cipher{MORUS640} and 256 for \cipher{MORUS1280}.

In the remainder, the $0$-th encryption step will denote the encryption step where our linear trail starts. Any encryption step could be chosen for that purpose, as long as at least four more encryption steps follow. In particular the $0$-th encryption step from the perspective of the trail does not have to be the first encryption step after initialization.

@@

\section{Symmetric Linear Combinations and \cipher{MiniMORUS}}
\label{sec:introminimorus}

To simplify the description of the attack, we assume all plaintext blocks are zero. However the same results would hold as long as the plaintext is fixed. Let $w$ denote the word size, i.e. $w = 32$ in the case of \cipher{MORUS640}, and $w = 64$ for \cipher{MORUS1280}. Recall that the inner state of the cipher consists of five $4w$-bit registers $S_0,\dots,S_4$, each containing four $w$-bit words.

@@

We begin with a few observations about the \StateUpdate{} function. Beside XOR and AND operations, the \StateUpdate{} function uses two types of bit rotations:
\begin{enumerate}
\item \emph{word-wise} rotations perform a circular shift on each word within a register;
\item \emph{register-wise} rotations perform a circular shift on a whole register.
\end{enumerate}
The second type of rotation always shifts registers by a multiple of the word size $w$. This amounts to a (circular) permutation of the words within the register: for example, if a register contains the words $(A, B, C, D)$, and a register-wise rotation by $w$ bits to the left is performed, then the register now contains the words $(B, C, D, A)$.

Since we want to build a linear trail, we are interested in linear combinations of bits within a register.
\begin{definition}
Recall that $w$ denotes the word size in bits, and $4w$ is the size of a register. A linear combination of the form:
\[
S^t_{i,j(0)} \oplus S^t_{i,j(1)} \oplus \dots \oplus S^t_{i,j(k)}
\]
is said to be \emph{symmetric} iff the set of bits $S^t_{i,j(0)}, \dots, S^t_{i,j(k)}$ is left invariant by a circular shift by $w$ bits; that is, iff:
\[
\{j(i) : i\leq k\} = \{j(i) + w \text{\rm{} mod } r : i\leq k\}.
\]
\end{definition}
\emph{Example.} The following linear combination is symmetric for \cipher{MORUS640}, i.e. $w = 32$:
\begin{equation}
S^t_{0,0} \oplus S^t_{0,32} \oplus S^t_{0,64} \oplus S^t_{0,96}.
\label{eq:symmetric}
\end{equation}

@@

Because the trail we are going to build is relatively complex, we will first describe it on \cipher{MiniMORUS}. We will then extend it to the full \cipher{MORUS} via the previous symmetry.

@@

\section{Linear Trail for \cipher{MiniMORUS}}

@@

The trail fragments are depicted on \Cref{fig:minimorusfragments}, and may be represented as follows (recall that the $b_i$'s are rotation constants):
\begin{alignat*}{2}
\alpha^t_i:\quad&& C^t_i &\to S^{t+1}_{0,i+b_0} \\
\beta^t_i:\quad&& C^t_i, S^t_{0,i} &\to S^t_{1,i} \\
\gamma^t_i:\quad&& S^t_{1,i}, S^{t+1}_{1,i+b_1} &\to S^t_{4,i} \\
\delta^t_i:\quad&& S^t_{4,i}, S^{t+1}_{4,i+b_4} &\to S^{t+1}_{2,i} \\
\varepsilon^t_i:\quad&& S^t_{2,i}, S^{t+1}_{2,i+b_2} &\to S^{t+1}_{0,i}.
\end{alignat*}

In all cases except $\alpha^t_i$, the trail fragment approximates a single AND gate by zero, which holds with probability 3/4, and hence the trail fragment has weight 1. In the case of $\alpha^t_i$, two AND gates are involved; however the two gates share an entry in common, so their sum can be written as:
\[
x \cdot y \oplus x \cdot z = x \cdot (y \oplus z).
\]
As a result, the trail fragment $\alpha^t_i$ also has a weight of 1. Another way to look at this phenomenon is that the trail holds for two different approximations of the AND gates: the alternative approximation is depicted by a dashed line on \Cref{fig:minimorusfragments}.

@@

\subsection{Trail Equation}
\label{sec:minitraileq}

@@

In order to build the equation for the full trail, we start with $E^3_0$:
\[
S^3_{2,0} \oplus S^{4}_{2,b_0} \oplus S^{4}_{0,0} = S^3_{3,0} \cdot S^t_{4,0}.
\]
In order to cancel the $S^{4}_{0,0}$ term on the left-hand side, we add in the equation $A^3_{-b_0}$ (where the sum of two equations of the form $a = b$ and $c = d$ is defined to be $a+c = b+d$). This yields:
\begin{align*}
&S^3_{2,0} \oplus S^{4}_{2,7} \oplus C^{3}_{0,-b_0}\\
=\; &S^3_{3,0} \cdot S^t_{4,0} \oplus S^t_{1,-b_0} \oplus S^t_{3,-b_0} \oplus S^3_{1,-b_0} \cdot S^3_{2,-b_0} \oplus S^3_{2,-b_0} \cdot S^t_{3,-b_0}.
\end{align*}
We then need to cancel two terms of the form $S^t_{2,i}$. To do this, we add in the equations $D^t_i$ for appropriate choices of $t$ and $i$. This replaces the two $S^t_{2,i}$ terms by four $S^t_{4,i}$ terms. By using equation $B^t_i$ four times, we can then replace these four $S^t_{4,i}$ terms by eight $S^t_{1,i}$ terms. By applying equation $B^t_i$ eight times, these eight $S^t_{1,i}$ terms can in turn be replaced by eight $S^t_{0,i}$ terms (and some ciphertext terms). Finally, applying $A^t_i$ eight times allows us to replace these eight $S^t_{0,i}$ terms by only ciphertext bits. Ultimately, for \cipher{MORUS640}, this yields the equation:

\begin{flalign*}
&&&C^0_{27} \oplus C^1_{0} \oplus C^1_{2} \oplus C^1_{8} \oplus C^1_{26} \oplus C^2_{1} \oplus C^2_{13} \oplus C^2_{15}&&\\
&&\oplus\; &C^2_{27} \oplus C^2_{31} \oplus C^3_{6} \oplus C^3_{12} \oplus C^3_{14} \oplus C^3_{20} \oplus C^4_{19}&&\\
&&=\; &S^0_{1,27} \oplus S^0_{3,27} \oplus S^0_{1,27} \cdot S^0_{2,27} \oplus S^0_{2,27} \cdot S^0_{3,27}&&\text{weight 1}\\
&&\oplus\; &S^1_{1,26} \oplus S^1_{3,26} \oplus S^1_{1,26} \cdot S^1_{2,26} \oplus S^1_{2,26} \cdot S^1_{3,26}&&\text{weight 1}\\
&&\oplus\; &S^1_{1,8} \oplus S^1_{3,8} \oplus S^1_{1,8} \cdot S^1_{2,8} \oplus S^1_{2,8} \cdot S^1_{3,8}&&\text{weight 1}\\
&&\oplus\; &S^2_{3,1} \oplus S^2_{1,1} \oplus S^2_{1,1} \cdot S^2_{2,1} \oplus S^2_{2,1} \cdot S^2_{3,1}&&\text{weight 1}\\
&&\oplus\; &S^2_{1,7} \oplus S^2_{3,7} \oplus S^2_{1,7} \cdot S^2_{2,7} \oplus S^2_{2,7} \cdot S^2_{3,7}&&\text{weight 1}\\
&&\oplus\; &S^2_{1,15} \oplus S^2_{3,15} \oplus S^2_{1,15} \cdot S^2_{2,15} \oplus S^2_{2,15} \cdot S^2_{3,15}&&\text{weight 1}\\
&&\oplus\; &S^2_{1,27} \oplus S^2_{3,27} \oplus S^2_{1,27} \cdot S^2_{2,27} \oplus S^2_{2,27} \cdot S^2_{3,27}&&\text{weight 1}\\
&&\oplus\; &S^1_{1,2} \oplus S^1_{3,2} \oplus S^1_{1,2} \cdot S^1_{2,2} \oplus S^1_{2,2} \cdot S^1_{3,2}&&\text{weight 1}\\
&&\oplus\; &S^3_{3,14} \oplus S^3_{1,14} \oplus S^3_{1,14} \cdot S^3_{2,14} \oplus S^3_{2,14} \cdot S^3_{3,14}&&\text{weight 1}\\
&&\oplus\; &S^2_{0,0} \cdot S^2_{1,0}&&\text{weight 1}\\
&&\oplus\; &S^2_{2,31} \cdot S^2_{3,31}&&\text{weight 1}\\
&&\oplus\; &S^2_{3,0} \cdot S^2_{4,0}&&\text{weight 1}\\
&&\oplus\; &S^3_{0,7} \cdot S^3_{1,7}&&\text{weight 1}\\
&&\oplus\; &S^3_{2,6} \cdot S^3_{3,6}&&\text{weight 1}\\
&&\oplus\; &S^3_{2,12} \cdot S^3_{3,12}&&\text{weight 1}\\
&&\oplus\; &S^4_{2,19} \cdot S^4_{3,19}.&&\text{weight 1}
\end{flalign*}

The trail for \cipher{MORUS1280} is very similar, and is given in \Cref{sec:traileq}.

\subsection{Bias of the Trail}
\label{sec:minibias}

In the equation for \cipher{MORUS640} from the previous section, each line on the right-hand side of the equality involves distinct terms, and each line has a weight of 1. By the piling-up lemma, it follows that if we assume distinct $S^t_{i,j}$ terms to be uniform and independent, then the expression on the right-hand side has a weight of 16. Hence the linear combination of ciphertext bits on the left-hand side has a bias of $2^{-16}$. The same holds for \cipher{MORUS1280} (cf. \Cref{sec:traileq}).

@@

To confirm that our analysis is correct, we ran experiments on an implementation of \cipher{MiniMORUS640} and \cipher{MiniMORUS1280}. In both cases, we measured the bias of two halves $\chi_1$ and $\chi_2$ of the full trail, as well as the bias of the full trail itself (denoted by $\chi$). Results are displayed on \Cref{tab:miniapproximations}. 
While our analysis from the previous section predicts a bias of $2^{-16}$, experiments confirm a slightly higher bias of $2^{-15.5}$ for \cipher{MORUS640}. The discrepancy of $2^{-0.5}$ probably arises from the fact that register bits across different rounds are not completely independent.

@@

\section{Trail for Full \cipher{MORUS}}

\subsection{Symmetrizing the Trail}
In the previous section, we have exhibited a linear trail on the reduced ciphers \cipher{MiniMORUS640} and \cipher{MiniMORUS1280}. We now turn to the full ciphers \cipher{MORUS640} and \cipher{MORUS1280}. As explained in \Cref{sec:introminimorus}, translating a trail from one of the \cipher{MiniMORUS} ciphers to the corresponding full \cipher{MORUS} cipher just amounts to making the linear combinations symmetric---indeed, that was the point of introducing \cipher{MiniMORUS} in the first place. That is, for every $S^t_{i,j}$ (resp. $C^t_j$) term in every trail fragment and every equation, we \emph{symmetrize} the term by adding in the terms $S^t_{i,j+w}$, $S^t_{i,j+2w}$, $S^t_{i,j+3w}$ (resp. $C^t_{j+w}$, $C^t_{j+2w}$, $C^t_{j+3w}$), where as usual $w$ denotes the word size. For example, if $w=32$ (for \cipher{MORUS640}), the term $S^3_{2,0}$ is symmetrized into:
\[
S^3_{2,0} \oplus S^3_{2,32} \oplus S^3_{2,64} \oplus S^3_{2,96}.
\]

In this way, the biased linear combination on \cipher{MiniMORUS640} given in \Cref{sec:minitraileq}, namely:
\begin{align*}
&C^0_{27} \oplus C^1_{0} \oplus C^1_{2} \oplus C^1_{8} \oplus C^1_{26} \oplus C^2_{1} \oplus C^2_{13} \oplus C^2_{15}\\
\oplus\; &C^2_{27} \oplus C^2_{31} \oplus C^3_{6} \oplus C^3_{12} \oplus C^3_{14} \oplus C^3_{20} \oplus C^4_{19}
\end{align*}
yields a biased linear combination on the full \cipher{MORUS640}, namely:
\begin{align*}
&C^0_{27} \oplus C^0_{59} \oplus C^0_{91} \oplus C^0_{123} \oplus C^1_{0} \oplus C^1_{2} \oplus C^1_{8} \oplus C^1_{26} \oplus C^1_{32} \oplus C^1_{34}\\
\oplus\; & C^1_{40} \oplus C^1_{58} \oplus C^1_{64} \oplus C^1_{66} \oplus C^1_{72} \oplus C^1_{90} \oplus C^1_{96} \oplus C^1_{98} \oplus C^1_{104} \oplus C^1_{122}\\
\oplus\; & C^2_{1} \oplus C^2_{13} \oplus C^2_{15} \oplus C^2_{27} \oplus C^2_{31} \oplus C^2_{33} \oplus C^2_{45} \oplus C^2_{47} \oplus C^2_{59} \oplus C^2_{63}\\
\oplus\; & C^2_{65} \oplus C^2_{77} \oplus C^2_{79} \oplus C^2_{91} \oplus C^2_{95} \oplus C^2_{97} \oplus C^2_{109} \oplus C^2_{111} \oplus C^2_{123} \oplus C^2_{127}\\
\oplus\; & C^3_{6} \oplus C^3_{12} \oplus C^3_{14} \oplus C^3_{20} \oplus C^3_{38} \oplus C^3_{44} \oplus C^3_{46} \oplus C^3_{52} \oplus C^3_{70} \oplus C^3_{76}\\
\oplus\; & C^3_{78} \oplus C^3_{84} \oplus C^3_{102} \oplus C^3_{108} \oplus C^3_{110} \oplus C^3_{116} \oplus C^4_{19} \oplus C^4_{51} \oplus C^4_{83} \oplus C^4_{115}.
\end{align*}
We refer the reader to \Cref{sec:traileq} for the corresponding equation on \cipher{MORUS1280}.

\subsection{Bias of the Full Trail}

Due to the observations from \autoref{sec:introminimorus}, the symmetrized trail on full \cipher{MORUS} more or less consists of four copies of the original trail on \cipher{MiniMORUS}. As such, one may expect that the weight of the symmetrized trail should simply be four times the weight of the corresponding \cipher{MiniMORUS} trail, namely 64 for both \cipher{MORUS640} and \cipher{MORUS1280}. However, reality is a little more complex, as the symmetrized trail does not simply behave as four copies of the original trail.

To understand why that might be the case, assume a nonlinear term $S^0_{2,0} \cdot S^0_{3,0}$ arises from some part of the trail, and another term $S^0_{2,0} \cdot S^0_{3,w}$ arises from a different part of the trail (where $w$ denotes the word size). Then when we XOR the various trail fragments together, in \cipher{MiniMORUS} these two terms are actually equal and will cancel out, since register-wise rotations by multiple of $w$ bits are ignored. However in the real \cipher{MORUS} these terms are of course distinct and do not cancel out.

In the actual trail for (either version of) full \cipher{MORUS}, this exact situation occurs when combining trail fragments $\beta^t_i$ and $\gamma^t_i$. Indeed, $\beta^t_i$ involves approximating the term $S^t_{2,i} \cdot S^t_{3,i}$, while $\gamma^t_i$ requires approximating the term $S^t_{2,i} \cdot S^t{3,i-w}$ (cf. \Cref{fig:trailcollision}). While in \cipher{MiniMORUS}, these terms cancel out, in the full \cipher{MORUS}, when adding all symmetric copies of the trail, we end up with the sum:
\begin{align}
&S^t_{2,i} \cdot S^t_{3,i} \oplus S^t_{3,i} \cdot S^t_{2,i+w}
\oplus S^t_{2,i+w} \cdot S^t_{3,i+w} \oplus S^t_{3,i+w} \cdot S^t_{2,i+2w}\notag\\
\oplus\; &S^t_{2,i+2w} \cdot S^t_{3,i+2w} \oplus S^t_{3,i+2w} \cdot S^t_{2,i+3w}
\oplus S^t_{2,i+3w} \cdot S^t_{3,i+3w} \oplus S^t_{3,i+3w} \cdot S^t_{2,i}.\label{eq:8circle}
\end{align}
It may be observed that the products occuring in the equation above involve 8 terms in a circular ring. The weight of this expression can be computed by brute force, and is equal to $3$.

\begin{figure}
  \substatesfalse
  % \substatesfalse to label state words and/or masks
  \centering
  \begin{subfigure}{.32\textwidth}
  \centering
  \begin{tikzpicture}[xscale=0.75,yscale=1.5]%{{{
    \printstate
    \draw[trail, beta]
      (C) -- node[right] {$i$} (lll-1)
      (lll-1) -- (tlll-1) (lll-1) -- (xor-1) (xor-1) -- (xnd-1)
      (tlll-1) -- (W-20) node[above] {$i$}
      (xor-1) -- (txor-1) %(txor-1) -- (W-21) node[above] {$i$}
      %(xnd-1) -- (and-1)
      (W40) node[below] {\phantom{$i$}}
      ;
    \draw[trail, gamma]
      %(W-21) node[above] {$i$}
      (txor-1) -- (xnd1)
      (xnd1) -- (xor1) (xor1) -- (lll1)
      %(xnd1) -- (and1)
      (xor1) -- (txor1)
      (txor1) -- (W-24) node[above] {$i$}
      (lll1) -- (W41) node[below] {$i+b_1$}
      ;
    %\draw[trail, beta,  dotted] (and-1) -- (and-1-|and1);
    %\draw[trail, gamma, dotted] (and1)  -- (and-1-|and1) node[above, black] {=};
    \draw[trail, beta]  (xnd-1) -- ++(-.5,0) coordinate (tmp) -- (tmp|-xnd1);
    \draw[trail, gamma] (tmp|-xnd1) node[below, black] {=} -- (xnd1);
  \end{tikzpicture}%}}}
  \caption*{\cipher{MiniMORUS}: weight 0 (not 2)} %$S^{j,1}_i, S^{j,4}_i, S^{j+1,1}_{i-1}$ ($w\!=\!1$)}
  \end{subfigure}
  \qquad
  \begin{subfigure}{.32\textwidth}
  \centering
  \begin{tikzpicture}[xscale=0.75,yscale=1.5]%{{{
    \printstate
    \draw[trail, beta]
      (C) -- node[right] {$i$} (lll-1)
      (lll-1) -- (tlll-1) (lll-1) -- (xor-1) (xor-1) -- (xnd-1)
      (tlll-1) -- (W-20) node[above] {$i$}
      (xor-1) -- (txor-1) %(txor-1) -- (W-21) node[above] {$i$}
      (xnd-1) -- (and-1)
      (W40) node[below] {\phantom{$i$}}
      ;
    \draw[trail, gamma]
      %(W-21) node[above] {$i$}
      (txor-1) -- (xnd1)
      (xnd1) -- (xor1) (xor1) -- (lll1)
      (xnd1) -- (and1)
      (xor1) -- (txor1)
      (txor1) -- (W-24) node[above] {$i$}
      (lll1) -- (W41) node[below] {$i+b_1$}
      ;
    \draw[trail, beta,  dashed] (and-1.east|-tanB-1) -- (tanB-1);
    \draw[trail, gamma, dashed] (tanB-1)  -- (tanB1) -- (tanB1-|and1.east);
  \end{tikzpicture}%}}}
  \caption*{\cipher{MORUS}: weight $4 \times 1$ (not $4 \times 2$)} %$S^{j,1}_i, S^{j,4}_i, S^{j+1,1}_{i-1}$ ($w\!=\!1$)}
  \end{subfigure}
  \caption{Weight of $\beta^t_i \oplus \gamma^t_i$ for \cipher{MiniMORUS} and \cipher{MORUS}.}
  \label{fig:trailcollision}
\end{figure}

In the end, for \cipher{MORUS1280}, since the trail fragment $\gamma^t_i$ is used four times, the phenomenon we just described adds a contribution of $4 \cdot 3 = 12$ to the overall weight of the full trail. This results in a total weight of $4 \cdot 16 + 12 = 76$ (where 16 is the weight of the trail on \cipher{MiniMORUS1280}). Due to the complexity of the trail, we have confirmed that this bias is correct by explicitely computing the full trail equation in \Cref{sec:traileq}, and computing its exact weight like we did for \cipher{MiniMORUS} in \Cref{sec:minibias}. That is, since the equation is quadratic, we may view it as a graph, which we split into connected components; we then compute the bias of each connected component separately by brute force, and then add up the weights of all components per the piling-up lemma. In the end, the full trail equation does confirm a weight of 76 for the full trail on \cipher{MORUS1280}.
